The sweet spot for local polygenic adaptation: a case study motivated by stickleback

### 2018-07-26 Stickleback_Paper.tex
more theory

\paragraph{New mutations}
The total rate of appearance of new mutations per lake is $\mu_L = 0.04$,
which are divided evenly in seven categories: neutral, and then additive, dominant, and recessive
in either direction.
This implies that a new additive or dominant effect mutation appears once every 87.5 generations,
on average.
Effect sizes are Exponential with mean $1/2$, and as above the probability
and so the probability that a dominant mutation manages to establish in a population differing from
the optimum by $z$ is roughly
$\int_0^\infty 4 \beta z u \exp(-2u) du = \beta z$,
and so the rate of establishment of dominant mutations is $\beta z / 87.5$,
i.e., about one such mutation every $2461/z$ generations.
The distribution of these successfully established mutations 
has density proportional to $u \exp(-2u)$, i.e., is Gamma with mean 1 and shape parameter 2.
Since additive alleles have half the effect in heterozygotes,
they have half the probability of establishment.
Combining these, we expect during the initial phase of adaptation,
when the populations begin at around distance $z=10$ from the optimum,
adaptive alleles to appear through mutation on a time scale of 250 generations,
and each to move the trait by a distance of order 1.

\paragraph{Migration}
A allele that moves the trait $z=1$ units in the freshwater direction
has fitness roughly $\exp(-\beta) \approx 1 - \beta$ in the marine environment
(which is close to optimal).
The product of population size and fitness differential in the marine environment
is therefore $2Ns = 8.9$, implying that these alleles are strongly selected against
but might occasionally drift to moderate frequency.

TODO: estimate rate at which migrant alleles move between lakes

>>>>

\paragraph{New mutations}
The total rate of appearance of new mutations per lake is $\mu_L = 0.04$,
which are divided evenly in seven categories: neutral, and then additive, dominant, and recessive
in either direction.
This implies that a new additive or dominant effect mutation appears once every 87.5 generations,
on average.
Effect sizes are randomly drawn from an Exponential distribution with mean $1/2$, 
and so the probability that a dominant mutation manages to establish in a population differing from
the optimum by $z$ is roughly
$\int_0^\infty 4 \beta z u \exp(-2u) du = \beta z$,
and so the rate of establishment of dominant mutations is $\beta z / 87.5$,
i.e., about one such mutation every $2461/z$ generations.
The distribution of these successfully established mutations 
has density proportional to $u \exp(-2u)$, i.e., is Gamma with mean 1 and shape parameter 2.
Since additive alleles have half the effect in heterozygotes,
they have half the probability of establishment.
During the initial phase of adaptation,
the populations begin at around distance $z=10$ from the optimum.
Combining these facts, we expect 
adaptive alleles to appear through mutation at first on a time scale of 250 generations,
with the time between local fixation of new alleles increasing as adaptation progresses,
and each to move the trait by a distance of order 1.

\paragraph{Standing variation}
An allele that moves the trait $z$ units in the freshwater direction in heterozygotes
has fitness roughly $\exp(-\beta z^2) \approx 1 - \beta z^2$ in the marine environment
(which is close to optimal).
The product of population size and fitness differential in the marine environment
for a mutation with $z=1$
is therefore $2Ns = 8.9$, implying that these alleles are strongly selected against
but might occasionally drift to moderate frequency.
The average frequency of such an allele in the marine environment at migration-selection equilibrium
is equal to the total influx of alleles per generation divided by the selective disadvantage,
which if $M = 2000 m$ is the number of immigrants per generation,
is $2 M / \beta z^2$. 
With $z=1$, the factor multiplying $M$ is $2/\beta z^2 \approx 1/200$:
since lakes have 400 genomes, 
as long as $M \ge 1$, the chances are good that any particular lake-adapted allele
that is present in all pre-existing lakes
will appear at least once in the fish that colonize a new lake.
However, an allele with $z=1$ only has probability of around 1/20 of establishing locally,
suggesting that we'd need $M \ge 10$ to ensure enough pre-existing genetic variation
that adaptation would happen entirely using the initial set of colonizers.
This corresponds to our highest two migration rates,
as in e.g., Figure \ref{fig:phenotype_ts2}B.

\paragraph{Migration}
If sufficient genetic variation is not present in a new lake initially,
it must appear by new mutation or migration.
Since a proportion $m$ of each lake is composed of migrants each generation,
it takes $1/m$ generations until the genetic variation introduced by migrants
equals the amount initially present at colonization.
This implies a dichotomy: either 
(a) adaptation is possible using variants present at colonization
or arriving shortly thereafter, or 
(b) adaptation takes many multiples of $1/m$ generations.

These calculations depend on there being no bottleneck in colonization of the lake;
if there is a bottleneck, then an additional factor must be added.

At what point do we expect new mutation to be more important than migration for adaptation?
By the calculations above, if $M \ge 10$, we expect initial diversity in a lake to be sufficient 
for adaptation, corresponding to our third-highest migration rate.
If this does not happen, then we expect adaptation to take a multiple of $1/m$ generations;
with our values, $1/m$ ranges from 20,000 to 20 generations.
Above we estimated that adaptive alleles due to new mutation fix locally about every 2,000 generations,
suggesting that at our second-lowest migration rate (where $1/m = 2,000$),
the two contributions of migration and new mutation are roughly equal.
This is in fact what we see: in Figure \ref{fig:MPFAI},
we see that at the second migration rate, alleles start to be shared between lakes,
while by the third migration rate, they are almost entirely shared.
